Demes: a specification for interchanging complex demography

### 2022-03-15 paper.tex
additional changes and showcase figure

\usepackage{filecontents,catchfile}
\begin{filecontents*}{software-table.tex}
\end{filecontents*}

@@

Here we present ``Demes'', a data model and file format specification for
complex demographic models developed by the
PopSim Consortium~\citep{adrion2020community}.
The Demes data model precisely defines the sizes and relationships
of populations, and it provides a way to explicitly encode the
information relevant to demography while avoiding repetition.
This data model is implemented in the widely used YAML
format~\citep{ben2009yaml}, which provides a good balance between human
and machine readability.
The specification precisely defines the required behaviour of implementations,
ensuring that there is no ambiguity of interpretation, and includes both a
reference implementation and an extensive suite of test examples and their
expected output.
The initial software ecosystem includes high-quality Python
and C parser implementations, as well as utilities for verification and
visualisation of Demes models, and has been implemented in several popular
inference and simulation methods (Table~\ref{tab-software}).
We hope that this data model and file format will be widely adopted
by the community, such that users can expect to simulate directly
from inferred models with little or no programming effort.

@@

Below we provide a brief overview of the population genetics models
that Demes supports and the components of the Demes infrastructure.
We then highlight an example usage, including visualization and simulation
of a multi-population demographic model.
Detailed descriptions of the population genetic model, parsers, scope, and
additional examples are given in the Appendix.

\CatchFileDef{\softwaretable}{software-table.tex}{}
\renewcommand{\arraystretch}{1.5}
\begin{table}
\begin{center}
\begin{tabular}{lp{12cm}}
\softwaretable
\end{tabular}
\end{center}
\caption{\label{tab-software}
Software support for Demes. We have included software infrastructure developed
for working with Demes models (parsing, validation, visualisation, etc)
as well as all known downstream software that implements the specification,
at the time of writing.}
\end{table}

\subsection*{Population genetic model}
1 paragraph

\subsection*{High level specification}
1 paragraph

\subsection*{Parsers}
and languages. 1 paragraph

@@

\begin{itemize}
    \item Static, instead of parameterized models (1 paragraph)
    \item Population level instead of genomic features (1 paragraph)
\end{itemize}

@@

@@

Stable and healthy software ecosystems require standard interchange
formats, allowing for the development of high-quality and long-lasting
tools that produce and consume the standard.
Demographic models are a key part of population genetics research,
and to date the transfer of inferred models to downstream simulations
has been ad-hoc, and conversions between the many different ways
of expressing such models is both labour intensive and fraught with errors.
The proposed Demes standard is an attempt to bridge this gap
between inference and simulation, and also to provide the foundations
for a sustainable ecosystem of tools built around this data model.
Table~\ref{tab-software} shows some initial infrastructure that we have
built as part of developing Demes, but many other useful tools
can be envisaged that consume, transform, or produce this format.

@@

\begin{figure}
\begin{minipage}{0.44\textwidth}
\begin{tcolorbox}
\inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
\end{tcolorbox}
\begin{tcolorbox}
\includegraphics[width=\textwidth]{fig/IM}
\end{tcolorbox}
\end{minipage}\hfill
\begin{minipage}{0.54\textwidth}
\begin{tcolorbox}
\inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
\end{tcolorbox}
\end{minipage}\\
\caption{\label{fig-example}
Example isolation-with-migration Demes model. (A) The compact non-redundant
representation expressed using YAML. (B) The same model in the fully-qualified
form. (C) A visual representation of the model using \texttt{demesdraw}.
The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
but are much more human-readable.
}
\end{figure}

@@

>>>>

\usepackage{catchfile}
\begin{filecontents*}{software-table.tex}
\end{filecontents*}

@@

Here we present ``Demes'', a data model and file format specification for
complex demographic models developed by the
PopSim Consortium~\citep{adrion2020community}.
The Demes data model precisely defines the sizes and relationships
of populations, and it provides a way to explicitly encode the
information relevant to demography while avoiding repetition.
This data model is implemented in the widely used YAML
format~\citep{ben2009yaml}, which provides a good balance between human
and machine readability.
The specification precisely defines the required behaviour of implementations,
ensuring that there is no ambiguity of interpretation, and includes both a
reference implementation and an extensive suite of test examples and their
expected output.
The initial software ecosystem includes high-quality Python
and C parser implementations, as well as utilities for verification and
visualisation of Demes models, and has been implemented in several popular
inference and simulation methods (Table~\ref{tab:software}).
We hope that this data model and file format will be widely adopted
by the community, such that users can expect to simulate directly
from inferred models with little or no programming effort.

@@

Below we provide a brief overview of the population genetics models
that Demes supports and the components of the Demes infrastructure.
We then highlight an example usage, including visualization and simulation
of a multi-population demographic model.
Detailed descriptions of the population genetics model, parsers, scope, and
additional examples are given in the Appendix.

\CatchFileDef{\softwaretable}{software-table.tex}{}
\renewcommand{\arraystretch}{1.5}
\begin{table}
    \begin{center}
        \begin{tabular}{lp{12cm}}
            \softwaretable
        \end{tabular}
    \end{center}
    \caption{
        \label{tab:software}
        \textbf{Software support for Demes.}
        We have included software infrastructure developed
        for working with Demes models (parsing, validation,
        visualisation, etc) as well as all known downstream 
        software that implements the specification,
        at the time of writing.
    }
\end{table}

\aprcomment{I want to keep the subsections that follow short and punchy.
Refer readers to the Appendix for details.}

\subsection*{Population genetics model}

Demographic models consist of one or more populations (or ``demes'') defined
by their size histories and the time intervals of their existence.
Individuals can move between populations based on their ancestor-descendant
relationships (i.e., the topology) or by continuous or discrete migration events.
The deomgraphic model itself omits information about genome biology, such as
genome size and architecture, ploidy, and mutation and recombination rates.
It also omits any information about selection operating within populations.
Within populations, allele-frequency dynamics are implicitly assumed to
follow the Wright-Fisher model, so that generations are discrete and
non-overlapping and population sizes are independent of mean fitness
(i.e., ``soft'' selection~\citep{christiansen1975hard}).
While these restrictions preclude some population-genetic scenarios of
interest, such as explicit spatial structure or
\aprcomment{stochastic carrying-capacity models, or something like that},
many current forwards- and backwards-in-time simulation software
make these same assumptions, so that Demes is relevant to a range of
widely-used simulators.
In Appendix~\ref{sec:appendix-pop-gen-model}, we expand on these concepts
and provide more details on the population genetics models supported by Demes.

\subsection*{High-level specification}

The high-level description of Demes models is focused on efficient human
understanding and avoiding errors.
We have adopted the widely used YAML format~\citep{ben2009yaml} as the
primary interface for writing and interchangin demographic models
(see Appendix~\ref{sec:appendix-spec} for details about our choice of
YAML formatting for demographic models).
Demographic models written in YAML format provide information about global
features of the model (such as time units and generation times),
populations (as ``demes'') and their existence intervals (as ``epochs''),
and gene flow between populations
(as continuous ``migrations'' or instantaneous ``pulse'' events)
(e.g., Figures~\ref{fig:showcase}~and~\ref{fig:IM}).
The high-level Demes specification encourages human understanding by avoiding
redundancy in the description where possible and by providing a mechanism for
specifying default values that are inherited hierarchically.

\subsection*{Parsers}

While the high-level specification is designed for human readability and
conciseness, the underlying data model suitable for software implementation
is redundant and exhaustive. 
Translation from the high-level specification to the low-level data model
requires resolving heierarchically-defined default values and verifying
relationships between populations and the validity of specified parameter
values.
Because this translation and validation requires significant
programming effort, we define a standard software entity as part of the
specification to perform this task (the parser),
which is intended to be shared by programs that support Demes as an input.
By having relatively few high-quality Demes parsers available as libraries
(ideally one per programming language), we ensure consistency across simulation
software in their implementation of Demes-specified models.
We provide a reference implementation written in Python to resolve any
potential ambiguities and to provide a helpful template for other
implementations, as well as an extensive test suite of examples and the
expected outputs.
In addition, we have high-quality parser implementations in the Python and
C languages (published under liberal open source licences), providing a
solid foundation for the software ecosystem (Table~\ref{tab:software}).

\aprcomment{redundancies with Appendix - that could be expanded and reworked a bit}

@@

The Demes specification is static by design -- we wish to unambiguously
describe a demographic model with a concrete set of parameters.
This simplicity means that we cannot directly specify parameter distributions
or estimated confidence intervals for those parameters.
While it is not difficult to imagine extending the specification in ways that
would allow this (Appendix~\ref{sec:appendix-scope}), it is not immediately
clear that the benefits are worth the increased parser complexity.

Demes is also designed to interface with a large number of simulation and
inference methods, so we have restricted the specification to only
describe demographic features at the population level.
The underlying discrete demographic structure is the common thread connecting
many widely-used population genetic methods, so we have omitted many
features of the underlying genome biology (as described above, and in
Appendix~\ref{sec:appendix-scope}).

@@

\begin{figure}[ht!]
    \centering
    \includegraphics{fig/showcase}
    \caption{
        \textbf{Illustration and simulation using Demes.}
        \aprcomment{Extended snippet and Tennessen YAML in the Appendix.}
    }
    \label{fig:showcase}
\end{figure}

@@

Stable and healthy software ecosystems require standard interchange
formats, allowing for the development of high-quality and long-lasting
tools that produce and consume the standard.
Demographic models are a key part of population genetics research,
and to date the transfer of inferred models to downstream simulations
has been ad-hoc, and conversions between the many different ways
of expressing such models is both labour intensive and fraught with errors.
The proposed Demes standard is an attempt to bridge this gap
between inference and simulation, and also to provide the foundations
for a sustainable ecosystem of tools built around this data model.
Table~\ref{tab:software} shows some initial infrastructure that we have
built as part of developing Demes, but many other useful tools
can be envisaged that consume, transform, or produce this format.

@@

\begin{figure}
    \begin{minipage}{0.44\textwidth}
        \begin{tcolorbox}
            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
        \end{tcolorbox}
        \begin{tcolorbox}
            \includegraphics[width=\textwidth]{fig/IM}
        \end{tcolorbox}
    \end{minipage}\hfill
    \begin{minipage}{0.54\textwidth}
        \begin{tcolorbox}
            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
        \end{tcolorbox}
    \end{minipage}\\
    \caption{
        \label{fig:IM}
        Example isolation-with-migration Demes model. (A) The compact non-redundant
        representation expressed using YAML. (B) The same model in the fully-qualified
        form. (C) A visual representation of the model using \texttt{demesdraw}.
        The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
        but are much more human-readable.
    }
\end{figure}

@@

\aprcomment{Need some information about defaults}
